---
title: "Assignment 1 - linear models"
output:
  html_document:
    toc: TRUE
    Toc_float: TRUE
---

## Collect data.  

Collect data about US states: 

https://worldpopulationreview.com/states 
scroll down to their "State Listing Articles" click on articles of interest, many of them will have links to csv files you can download. For example: 
blob:https://worldpopulationreview.com/37b7dc59-9bc3-48f0-a4a3-8a2e2b000d6a

You need to collect data on at least 10 different attributes. Some could be continuous and some categorical (midwest). 


## Define research questions. 
Decide what you want to study, i.e. chose one (or a few) dependent variables and other that are independent. 

Build linear models, generate plots and make condifent statement. 

### Submissson. 
Please submit an html/pdf (i.e. preview or knitt your notebook). The code should be annotated with proper markup and comments in the text. Show figures for all key steps as needed. 

Load packages needed for analysis
```{r packages}
## load the packages necessary for your analysis



```

Read data
```{r dataimport}
## import the data for your analysis


```

Include any data cleaning steps
```{r datacleaning}


```

Exploratory figures 
```{r dataviz1}

```

Define research questions. 
* This is questions 1
* justrified by reason 2. 

What statistics do you need to do to
check assumptsions etc. 
```{r stat_testing}

```



