---
title: "Assignment 1 - linear models"
author: "Your name"
output:
  html_document:
    toc: TRUE
    Toc_float: TRUE
---

## Define research questions and collect data  

Have a look at the data available on https://worldpopulationreview.com/states 

Scroll down to their "**State Listing Articles**" click on articles of interest. Many of them will have links to csv files you can download. 

Decide **what you want to study**, i.e. chose one (or a few) dependent variables and other that are independent. 

You need to collect a few data sets, in order to reach about 10 different attributes (which would corresponds to 10 column in the final data set). Some could be continuous and some categorical. 

## Explore and analyse the data, and interpret the results

Build linear models, generate plots and make confidence statement. 

## Submission

You can work on RStudio Cloud, or on the desktop version of RStudio.

Please submit an **html/pdf** (i.e. preview or knitt your notebook). The code should be **annotated** with proper markup and comments in the text. Show figures for all key steps as needed. 

###

Load packages needed for analysis
Note - If you are not sure which package you need now, you can come back to this section later to update it as needed.

```{r packages}

## load the packages necessary for your analysis



```

Read data
Note - If you are working on RStudioCloud, you can put all the csv files in a zip folder, and upload this zip from the 'Files' windows of the RStudio interface.

```{r dataimport}

## Import the data for your analysis


```

Data cleaning and merging 

```{r datacleaning}


```

Exploratory figures 

```{r dataviz1}



```

Define research questions and hypothesis
* Question/hypothesis: 
* Justification: 

What statistics do you need to get to
Do not forget to check that model assumptions are verified

```{r stat_testing}



```

How do you interpret the results?
