---
title: "Module 1.3 Plotting"
author: "A. Gamble, I. Ali, N. Zaitlen, R. Wollman"
output: 
 html_document:
 toc: TRUE
 toc_float: TRUE
---

## Intro to plotting

One of the most powerful features of R is its functionality with data
visualization and plotting. Various developers in the R community have created
robust tools for data visualization and plotting. Some plotting libraries are
immensely extensive, allowing you to create nearly any kind of plot you would
like and more. 

The R developers community has created a variety of packages that are useful
for plotting. Base R has a highly customizable plotting function and is
used on occasion (see plot()). Searching forums is a great way to learn
about the ways you can customize features needed for your own data visualization
approaches. See: 
[r-bloggers.com/some-basics-for-base-graphics](https://www.r-bloggers.com/some-basics-for-base-graphics/)
More recently, other plotting libraries have been developed and implemented
in various contexts (`plotly`, `leaflet`, `pheatmap`, etc). 

The most commonly used visualization tool in R is `ggplot2` package which has immense 
customizability and the most extensive documentation and instructional resources 
of the plotting libraries. ggplot2 is based on what is called the 'grammar of graphics'.
This is the idea that you can build nearly every graph from the same components: 
(1) a data set, (2) a coordinate system and (3) 'geoms' which are visual 
marks that represent data points. ggplot2 is able to create a diverse set 
of plots using these three characteristics. Here we will explore how to 
construct a few example plots using ggplot2. For a more exhaustive list 
of ggplot2 functions please review the [ggplot2 cheat sheet](https://hbctraining.github.io/Intro-to-R-flipped/cheatsheets/data-visualization-2.1.pdf). 

It is worth fully exploring ggplot2 documentation. There are many 
customizable features for each geometry making ggplot2 one of the most 
comprehensive packages for data visualization in R. 

The art of data visualization requires both an understanding of the data that
you are working with and also an awareness of the important features of the 
data that you want to highlight. Experimenting with different visualization 
strategies can help you better pin down the most informative ways of displaying 
your data. When working with your own data it can be important to look at mean
or median values, the variance and spread, correlations and clusters.

Through our exercises in this session, we aim to highlight some of the more
common plots you might come across in your research. We will cover how to make
these plots using ggplot, or other relevant plotting libraries

## Setup

```{r setup, include = FALSE}

knitr::opts_chunk$set(echo = TRUE)
library(readr) # read data from the web
library(ggplot2) # plotting library
library(ggExtra) # extra features for plotting library
library(dplyr) # data transformation
library(reshape2) # data transformation
library(RColorBrewer) # better color controls for plotting

```

## Data import

Using the tools and strategies from Module1.1 and Module1.2 we created a data set
to simulate a time-course gene expression experiment. Using read.csv() import the
data to create an object called 'dummydata'.

In this data set we will observe the expression of 15 genes (measured in counts per
million) over 8 hours. We will use various plotting strategies to understand various
characteristics of this data set.

```{r dummydataimport}

# using the read.csv() command import the GeneExpressionData.csv file
dummydata <- read.csv("GeneExpressionData.csv", header = TRUE, sep = ",")

# notice that the data is in long form.
View(dummydata)

```

## Plotting one variable

**Histograms** are some of the most common plots you will see in your research. In this case, histograms measure the number of values that fall into pre-defined bins. We can use
`geom_histogram()` to look at the distribution of counts per-million for all the genes
in our data set. 

We can set a fixed number of bins across a data set, or we can set a fixed bin width to
visualize the data.

```{r histogram}

# Use geom_histogram to create a histogram to show the gene expression counts
# for all of the data

# All ggplot commands follow a similar syntax. First within ggplot() you define the
# data set, and the aesthetics (what you want to set as x, y, or z)
# For histograms you only need to include one variable within aes()

# Notice that you can add layers and features to ggplots with a + symbol

ggplot(dummydata, aes(MeanCounts)) + 
 geom_histogram(bins = 20)

# It looks like our data is a bit skewed
# Lets set the fill color to change with the gene names to see 
# if some genes have higher overall expression than others

ggplot(dummydata, aes(MeanCounts, fill = GeneNames)) + 
 geom_histogram(bins = 20)

# We can change the width of the bins as well

ggplot(dummydata, aes(MeanCounts, fill = GeneNames)) + 
 geom_histogram(binwidth = 10)

# Pro tip: you can assign ggplots as objects in your global environment and call them
# later to further organize your data visualizaitons

```

**Dot plots** are a bit less common, but they can be used with similar results
as a histogram. 

```{r dotplot}

# Use geom_dotplot() with a defined dotsize, binwidth and method to produce
# a plot similar to our original histogram

ggplot(dummydata, aes(MeanCounts)) + 
 geom_dotplot(dotsize = 0.5, binwidth = 10, method = "histodot")

# Notice the y axis - What is it showing us?

```

## Adding lines to plots

Using our example we can take a look at how gene expression changes over time 
using line plots. There are multiple kinds of line plots that you can create with
ggplot(). `geom_line()` uses quantitative x and y aesthetics to connect the dots for
related variables.

```{r plottinglines}

# Plot Mean Counts (y) over time (x). Use color = to distinguish between genes
ggplot(dummydata, aes(x = TimePoint, y = MeanCounts, color = GeneNames)) + 
 geom_line()

# Filter on one or more genes in order to view individual genes on the plot
# using the tidyverse package you can quickly filter on specific genes
# This will make it easier for us to view trends among genes

# create a data frame from the dummydata set that only contains gene 1 
# through gene 5
gene15plot <- dummydata %>%
 filter(GeneNames == c("Gene01", "Gene02", "Gene03", "Gene04", "Gene05"))

# Recreate our previous line plot for our filtered data set. Notice if we do not
# include "x = " or "y = " ggplot will assume you list x first and y second.
ggplot(gene15plot, aes(TimePoint, MeanCounts, color = GeneNames)) + 
 geom_line()

```

Add additional layers to emphasize features of the plot. Let's create a line plot
from genes 6 through 10 and add points to emphasize transition points. 

```{r addinglayers}

# Create a data frame from the dummydata set that only contains gene 6 
# through gene 10
gene610plot <- dummydata %>%
 filter(GeneNames == c("Gene06", "Gene07", "Gene08", "Gene09", "Gene10"))

# Add geom_point to add points to your line plot
ggplot(gene610plot, aes(x = TimePoint, y = MeanCounts, color = GeneNames)) + 
 geom_line() + 
 geom_point()

# Use geom_smooth() instead of geom_line() to create a smoothed line 
# of the data
ggplot(gene610plot, aes(x = TimePoint, y = MeanCounts, color = GeneNames)) + 
 geom_smooth() + 
 geom_point()

# Modify characteristics of geom_smooth() to improve data readability
# geom_smooth() creates a linear model based on the data points and
# displays a confidence interval
# You can set the method used to generate
# the linear model or remove the confidence interval to compare lines 
# with the 'se = " argument
ggplot(gene610plot, aes(TimePoint, MeanCounts, color = GeneNames)) + 
 geom_smooth(method = "auto", se = FALSE) + 
 geom_point()

```

Data visualization is always a balance between increasing data density while
maintaining readability. If you want to quickly separate plots based on specific
facets, use the `facet_wrap()` command and indicate what facets used for separating.

```{r plottinglinesfacets}

# Use geom_smooth() to look at how genes 6 through 10 change over time 
# and use the GeneNames variable as a facet to make it easier to see the
# the confidence intervals.

# As your plots increase in complexity, hit return after every + to auto
# indent and improve readability of your code

ggplot(gene610plot, aes(x = TimePoint, y = MeanCounts, color = GeneNames)) + 
 geom_smooth(method = "auto") + 
 geom_point() + 
 facet_wrap(facets = "GeneNames") 

```

## Boxplots and colors

Another extremely important aspect of data visualization includes **depicting sample ranges**, spread and variability. There are number of ways to do this
for our lesson we will cover **box plots** and **violin plots** as these are some 
of the most common used for plotting.

```{r boxplot}

# Create a boxplot showing the distribution of gene expression for genes 11-15
# across all time points
gene1115plot <- dummydata %>%
 filter(GeneNames == c("Gene11", "Gene12", "Gene13", "Gene14", "Gene15"))

ggplot(gene1115plot, aes(x = GeneNames, y = MeanCounts, fill = GeneNames)) + 
 geom_boxplot()

# Modify the color palette using RColorBrewer
# scale_fill_brewer() modifies the fill of plot objects
# scale_color_brewer() modifies the color of the plot objects
# select the palette from many options which you can find
# https://www.r-graph-gallery.com/38-rcolorbrewers-palettes.html

ggplot(gene1115plot, aes(x = GeneNames, y = MeanCounts, fill = GeneNames)) + 
 geom_boxplot() + 
 scale_fill_brewer(palette = "Blues")

# Customize Axis Titles and Labels. Notice that when you hit return after a
# comma, + or %>%; R will automatically indent to make the code more readable
ggplot(gene1115plot, aes(x = GeneNames, y = MeanCounts, fill = GeneNames)) + 
 geom_boxplot() + 
 scale_fill_brewer(palette = "Greens") + 
 labs(title = "Box Plots of Gene Counts", 
      x = "Gene IDs", 
      y = "Counts per Million")

```

`RColorBrewer` has a variety of preset color palettes and additional features to
expand and modify color palettes as data increases in complexity. As an entry
point, lets take a look at the distribution of gene expression for all genes
in our dummy data set.

## Violin plots

Violin plots are a means of depicting data distribution for multiple
categorical variables simultaneously. For this example. Let's use violin plots
to look at the range for each gene across all time points.

```{r violinplots}

# Use our dummydata set and geom_violin to create violin plots of our data
ggplot(dummydata, aes(x = GeneNames, y = MeanCounts, fill = GeneNames)) + 
 geom_violin(scale = "width") + 
 labs(title = "Violin Plots of Gene Counts",
      x = "Gene IDs", 
      y = "Counts per Million")

# The x axis is a bit crowded
# We can rotate the axis text by modifying the theme()
# Using theme() modifications you can customize nearly every
# feature of your ggplot

# Use labs() to add a plot Title and customize X and Y axis text
ggplot(dummydata, aes(x = GeneNames, y = MeanCounts, fill = GeneNames)) + 
 geom_violin(scale = "width") + 
 theme(axis.text.x = element_text(angle = 45, hjust = 1)) + 
 labs(title = "Violin Plots of Gene Counts", 
      x = "Gene IDs", 
      y = "Counts per Million")

# Lets add a dot plot on top of our violin to see how each individual point falls
# into the violin plot's depicted distribution
ggplot(dummydata, aes(x = GeneNames, y = MeanCounts, fill = GeneNames)) + 
 geom_violin(scale = "width") + 
 geom_dotplot(binaxis = "y", stackdir = "center", fill = "black", color = "white", dotsize = 0.7) + 
 theme(axis.text.x = element_text(angle = 45, hjust = 1)) + 
 labs(title = "Violin Plots of Gene Counts", 
      x = "Gene IDs", 
      y = "Counts per Million")

```


## Bar plots

It is important to note that R plots all the values that you give it. What happens if you want to create a bar plot that investigates the average gene expression for genes 1 through 5?

```{r barplot}

# Plot the values like we did with he violin and box plots
# What do you see?
ggplot(gene15plot, aes(x = GeneNames, y = MeanCounts, fill = GeneNames)) + 
 geom_col() + 
 scale_fill_brewer(palette = "Purples") + 
 labs(title = "Bar Plots of Average Gene Counts", 
      x = "Gene IDs", 
      y = "Counts per Million")

# For bar and column plots, you will need to independently calculate the mean/sd for
# each of the genes and then plot them as individual values
# Use the group_by() and summarize() functions to quickly do this

gene15summary <- gene15plot %>%
 group_by(GeneNames) %>% # the group_by() function works similarly to "by = " 
 summarise(Mean = mean(MeanCounts), 
 StanDev = sd(MeanCounts))

ggplot(gene15summary, aes(x = GeneNames, y = Mean, fill = GeneNames, 
                          ymin = Mean-StanDev, ymax = Mean + StanDev)) + 
 geom_col() + 
 scale_fill_brewer(palette = "Purples") + 
 geom_errorbar(width = 0.5) + 
 labs(title = "Bar Plots of Average Gene Counts",
      x = "Gene IDs", 
      y = "Counts per Million")

```

## Modifying colors with added complexity

```{r complexcolorscales}

# Most color palettes have a fixed number of default colors in their range
# In the case of RColorBrewer most color palettes have a range of 9 colors
# in their default setting
# In the case where we have more than 9 categories, we need to expand 
# the palette in order to properly color the data. See what happens when you use
# the default fill settings on our violin plots from earlier

ggplot(dummydata, aes(x = GeneNames, y = MeanCounts, fill = GeneNames)) + 
 geom_violin(scale = "width") + 
 geom_dotplot(binaxis = "y", stackdir = "center", fill = "black", color = "white", dotsize = 0.7) + 
 theme(axis.text.x = element_text(angle = 45, hjust = 1)) + 
 scale_fill_brewer(palette = "Blues") + 
 labs(title = "Violin Plots of Gene Counts", 
      x = "Gene IDs", 
      y = "Counts per Million")

# Notice that genes 10-15 have no color
# So in this case we need to expand the color palette # to match the number 
# of variables we are considering
# To do this we can count the number of  unique genes in our data set and
# tell RColorBrewer to expand the range to that many colors
ncolors <- length(unique(dummydata$GeneNames))

# Once we have the number of colors that we need. We can use the scale_fill_manual()
# command to create the color palette needed. Notice the syntax below
# Getting the parenthesis right is crucial for this code to work

ggplot(dummydata, aes(x = GeneNames, y = MeanCounts, fill = GeneNames)) + 
 geom_violin(scale = "width") + 
 scale_fill_manual(values = colorRampPalette(brewer.pal(9, "Blues"))(ncolors)) + 
 geom_dotplot(binaxis = "y", stackdir = "center", fill = "black", color = "white", dotsize = 0.7) + 
 theme(axis.text.x = element_text(angle = 45, hjust = 1)) + 
 labs(title = "Violin Plots of Gene Counts", 
      x = "Gene IDs", 
      y = "Counts per Million")

```

## Heatmap

```{r heatmap}

ggplot(dummydata, aes(x = as.character(TimePoint), y = GeneNames)) + 
 geom_tile(aes(fill = MeanCounts)) + 
 labs(title = "Heatmap of Average Gene Counts", 
      x = "Time Point", 
      y = "Counts per Million")

```

## Exporting figures

We can export publication-quality figures with `ggsave()`.

```{r save}

ggsave(
 "Heatmap.pdf", # do not forget the format
 plot = last_plot(), # default is last plot but we can use plot names too
 device = "pdf",
 width = 10,
 height = 10,
 units = c("cm"),
 dpi = 300
)

```
