---
title: "Module 2.1 - Correlations"
author: "A. Gamble, I. Ali, N. Zaitlen"
output: 
 html_document:
 toc: TRUE
 toc_float: TRUE
---

## Setup

In today's session, we will use the `ggplot2` package for plotting.

```{r setup, include = FALSE}

knitr::opts_chunk$set(echo = TRUE)

library(ggplot2) # plotting library

```

## Data exploration

### Loading and printing the data

For this example, we are going to use the built-in Iris data set.

```{r data}

head(iris)

```

Below, we explore potential associations between petal length and width, and between sepal length and width.

### Visualizing the data

```{r scatterplot1}

# plot() function to visualize the data with a scatter plot
plot(iris$Petal.Width, iris$Petal.Length, # Data
     pch = 19, # Point shape
     col = c("red", "green3", "blue")[iris$Species], # Color
     main = "Iris Data - Petals", # Title
     xlab = "Petal Width", ylab = "Petal Length" # Axis labels
     )

# legend() function to add a legend
# the arguments have to match the ones used to generate the plot
legend("topleft", # Position
       legend = c("Setosa", "Versicolor", "Virginica" ), # Labels, could also use levels(iris$Species)
       pch = 19, # Point shape
       col = c("red", "green3", "blue")) # Color

```

```{r histogram1}

# hist() function to visualize the data with an histogram
hist(iris$Petal.Length, # Data
     main = "Iris Data - Petals", # Title
     xlab = "Petal Length", ylab = "Frequency" # Axis labels
     )

```

## Testing correlations

### Pearson's correlation 

The Pearson's correlation is used to measure the strength and direction of association between two variables. The Pearson's correlation coefficient can be obtained using the `cor()` function by setting to method to `pearson`.

```{r pearson-cor}

# cor() function to calculate the correlation coefficient
cor(iris$Petal.Length, iris$Petal.Width, # Variables between which we want to measure the association
    method = c("pearson") # Method
    ) 

```

Correlation coefficients can also be obtained using the `cor.test()` function, which also returns a confidence interval of the correlation coefficient. 

```{r pearson-cor.test}

# cor.test() function to calculate the correlation coefficient
# and the associated confidence interval
cor.test(iris$Petal.Length, iris$Petal.Width, # Variables between which we want to measure the association
         method = "pearson" # Method
         )

```

### Spearman's rank correlation

The Spearman's rank correlation is a non-parametric equivalent to Pearson's correlation. It is used to measure the strength and direction of association between two ranked variables. The Spearman's rank correlation coefficient can also be obtained using the `cor()` function or the `cor.test()` function by setting to method to `spearman`.

```{r spearman}

# cor() function to calculate the correlation coefficient
cor(iris$Petal.Length, iris$Petal.Width, # Variables between which we want to measure the association
    method = c("spearman") # Method
    ) 

# cor.test() function to calculate the correlation coefficient
# and the associated confidence interval
cor.test(iris$Petal.Length, iris$Petal.Width, # Variables between which we want to measure the association
         method = "spearman" # Method
         )

```

Note that, as Spearman's method is based on ranks, it cannot compute exact p-value when there are tied values (i.e., when two are more observations are equal).

### Permutations

Add explanations...

We start by writing a function that calculates the p-value and correlation coefficient using permutations. 

```{r permutations-function}

my_perm_corr_test = function(x, y, iter = 1000) {
  r_perm = array(dim = iter)
  for (i in 1:iter) { # seq(iter) same as 1:iter
    r_perm[i] = cor(x, sample(y))
  }
  r_real = cor(x, y)
  r_perm_abs = abs(r_perm)
  r_real_abs = abs(r_real)
  bigger_thans = r_real_abs > r_perm_abs
  p_grater_thans = mean(bigger_thans)
  pval = 1 - p_grater_thans
  rtn = list("r_perm" = r_perm,
             "pval" = pval,
             "r_real" = r_real)
  return(rtn)
}

```

We can now apply our `my_perm_corr_test()` to the petal data. 

```{r permutations-petals}

# Execute the my_perm_corr_test() function
prmtst = my_perm_corr_test(iris$Petal.Length, iris$Petal.Width, iter = 1000)

# Visualize the distribution of the permutations
hist(prmtst$r_perm)

# Print the calculated p-value and correlation coefficient
prmtst$pval
prmtst$r_real 

```

We then apply our `my_perm_corr_test()` to the sepal data. 

```{r permutations-sepals}

# Execute the my_perm_corr_test() function
prmtst = my_perm_corr_test(iris$Petal.Length, iris$Petal.Width, iter = 1000)

# Visualize the distribution of the permutations
hist(prmtst$r_perm)

# Print the calculated p-value and correlation coefficient
prmtst$pval
prmtst$r_real 

```
