---
title: "Module 1.1 - Intro to R"
author: "I. Ali, A. Gamble, N. Zaitlen, R. Wollman"
output: 
 html_document:
 toc: TRUE
 toc_float: TRUE
---

## Overview

R is one of the most commonly used and powerful statistical programming languages. It is used to analyze and manipulate data for virtually every professional industry. Over the last 10 years, powerful new tools within R have emerged to make data manipulation and visualization an easier and more accessible process. R is an open source software which means that it will remain freely available to those who need to use it, but also may change its capabilities and improve over time. 

*How does R work?*

R reads your data (e.g., from a text or and Excel file) and makes an image of it, which is stored in temporary memory. When you edit your data, you are only editing an image of it; the original data set is unaffected. You can then create and alter the data sets within R, which can then be written back to a permanent file. This is called **nondestructive editing**. You can also create objects (like lists, vectors, matrices and arrays) directly within R, as opposed to reading them in from a file.

R can do almost anything you can think of doing with your data, which greatly expands the amount of creativity you can apply to your analyses. By using **R scripts** ('.R' files) or **Rmarkdown scripts** ('.Rmd' files, such as this one), you can keep a record of everything you have done to manipulate and analyze your data. Those benefit you later, and can be published or shared with colleagues. In particular, Rmarkdown scripts allow you to easily mix text, code snippets, and code output, generative what we call **literate code**, readable by both machines but also people.

## R basics - Data types and data structures

R works most efficiently with data in the form of **vectors**, which you can think of as series of info that fall under a particular **class**. When multiple vectors of the same length are joined up together they create a **data frame**, which might look similar to a table you would see in Excel. A data frame can have multiple classes of data. A **data matrix** looks like a data frame, with the exception that it uses only one class of data. An **array** allows you to work with multidimensional data. In an array, you can package individual values (in other words scalars) vectors, data frames and matrices. We will see various examples of these items over the course of the lesson.

R recognizes several classes of data that can be 'packaged' into vectors. The most common data types include characters, numerics, integers, logicals and factors. **Characters** are strings of alphanumeric characters that are written with quotation marks around them (e.g., `'this is a string of characters'`). Integers and numerics, are interpreted as different classifications of numbers, **Integers** represent whole numbers and **numerics** also include decimals and numbers in scientific notation. **Logicals** are TRUE / FALSE designations which are very useful for filtering, creating for loops, or if/else components in your code. **Factors** can be useful for creating groups or categories of data. We will see examples of each of these over the course of the lesson. 

To run code in an Rmarkdown script (like this one), you can highlight what code you would like to run and click the 'Run' button in the top right corner of this window pane. You can also use the Ctrl+Return shortcut.

Let's start by creating some simple objects and printing their values.

```{r my_chunck}

# Pro-tip: keep your code explicit by adding comments (text ignored by R) with '#'

# create an object by writing in the object name and assigning it with a value using either the = or the <- symbols

x = 5
y <- 1 # for creating objects, this is the preferred way since = symbols are used in other contexts

animal <- "cat"

isDog <- FALSE

# To view a value, you can either output its value by allowing R to print it...
x

# ... but if you have many outputs, they might not be printed out
# To make sure you print something, use print()
print(y)

# Pro-tip: when encountering a new function, you can use "?<FunctionName> to access R help
# E.g., ?print

# You can even combine different types of objects and free text by writing a character string string and pasting the variable alongside it
paste(x)
paste("X value is:", x)

```

### Exercise 1

R can function as a calculator. Create a new object z that has the value of 2*y-x and print it out.

```{r exercise1}

z <- 2*y-x

```

## R basics - Vectors and matrices #1

Vectors are the most fundamental data structure in R. Vectors are used to store a bunch of values in a series. Indexing is a term used to articulate the order/position of values within a data structure.

```{r vectors}

# Write a series of consecutive integers using a ':' between two numbers
1:4

# Creating a vector object called legs
legs <- c(4, 4, 8, 6) # note the c() function allows you to create a series of non-consecutive numbers, or a series of character strings

legs[1] # this will find the first item in the legs vector

legs[c(2, 4)] # this will find the second and fourth item in the legs vector

```

### Exercise 2

Print the value of the third item in legs with indexing.

```{r exercise2}

print(legs[3])

```

## R basics - Vectors and matrices #2

You can also give specific names to positions in a vector (and many other things) using the `names()` function.

```{r vectornames}

specie <- c("cat", "dog", "spider", "ant")
names(legs) <- specie

# When positions are assigned with names, the names can be searched with the indexing function
legs["spider"] 

```

### Exercise 3

Print the number of legs ants have.

```{r exercise3}

print(legs["ant"])

```

## R basics - Vectors and matrices #3

Operations between vectors are done in order, or by the corresponding element.

```{r vectormath}

# You can do math with numeric vectors of equal length
eyes <- c(2, 2, 8, 2)
eye_to_leg <- eyes/legs

# Calculate averages with the mean() function - see also median(), sd(), min(), max()
mean(eye_to_leg) 

# Math with vectors of different length will give you warnings. Try running the code below.
eyes/c(1, 3, 4)

```

### Exercise 4 

Create a new vector with an attribute of your choice (how scary they are?)
print out the attribute of spider from your vector

```{r exercise4}

scariness <- c(5, 5, 8, 6)
names(scariness) <- specie
scariness["spider"]

```

## R basics - Vectors and matrices #4

Accessing vector via multiple indexes or logical operations

```{r logicaloperations}

# We can see the value and name of items called by indexing
# Notice the names are retained if one of the two vectors
# that were part of a mathematical operation has assigned names

eye_to_leg[c(1, 2)]

# Lets create a vector of boolean values that describes if the animals are, or ar not, mammals
is_mammal <- c(TRUE, TRUE, FALSE, FALSE)

# Notice that boolean operators (TRUE/FALSE) are written in all-caps or as single letters (T/F)
# and they do not require quotations around them

# Using the is_mammal designation, we can call out all the values in a vector that are TRUE
eye_to_leg[is_mammal]

# Selectively run operations like mean() by calling only the values that are TRUE for a condition
mean(eye_to_leg[is_mammal])

# Output the indexes of items that are TRUE using which()
which(is_mammal)

# You can also create boolean vectors using logical operations to select for conditions
has_two_eyes <- eyes == 2

```

### Exercise 5

Create a new boolean vector more_than_four_legs and use it to print the names of species with >4 legs. Do it twice, once using the original legs vector and once using the names() function

```{r exercise5}

# create the vector
more_than_four_legs <- c(FALSE, FALSE, TRUE, TRUE)
more_than_four_legs <- c(F, F, T, T) # same as above
names(more_than_four_legs) <- specie

# Print the values
# Option 1
legs[more_than_four_legs]

# Option 2
specie[more_than_four_legs]

# Another view
legs > 4


```

## R basics - Vectors and matrices #5

Matrices are another common data structure. Think of them as vectors forced into rectangular shapes. Notice that matrices can only contain one data type. Numerics, Character, Logical, etc.

```{r matrices}

# Use the matrix() function to create a 3x3 numeric matrix using the numbers 1 through 9
# Search for the matrix() function under help to find the syntax for the command 
m <- matrix(1:9, byrow = TRUE, nrow = 3)
rownames(m) = c("t1", "t2", "t3") # assign names to rows
colnames(m) = c("p1", "p2", "p3") # assign names to columns

# Use the print() function to print the values into the console
print(m)

# You can use indexing to access elements of a matrix. 
# Use commas to indicate the dimensions
# What are the following three lines doing? Write as a note below each line.
m[2]
# call the second item
m[2, ]
# call the second row
m[, 2]
# call the second column

# Pro-tip: Use short but informative names when creating objects, so you can remember
# what you have made. Using m for matrix, can be confusing if you have to create more than
# one different matrix in your data

```

### Exercise 6

Print out the third item of the second row.

Print all the items that are p3.

```{r exercise6}

# third item of second row
m[2, 3]

# all items in column p3
m[, 3]

```

## R basics - Factors

Factors are another "core type" in R that is useful for stats. They often contain categorical variable information. They generally have a limited set of different values and are often used in statistics to ensure statistical modeling functions treat such data correctly.

```{r factors}

# Create a vector of character strings indicating the taxonomic class of our respective animals
class <- c("Mammal", "Mammal", "Arachnid", "Insecta")

# You can use the factor() function to assign character vectors as factors. You can also do this with the as.factor() function
class_factor <- factor(class)
levels(class_factor)

# The by() function by applies a function to groups of elements in vectors (and other data
# structures) by factor. In our example below there are three arguments: (1) the vector of interest, 
# (2) the vector of factor values, and (3) the operation to conduct (in this case, mean)
by(eye_to_leg, class_factor, mean)

# This should return average eye-to-leg ratio for each class in our data set

```

### Exercise 7

Using the approach from above, create new factor with two levels "big" and "small" and calculate the maximum number of legs per group.

```{r exercise7}

animalsize <- c("big", "big", "small", "small")

size_factor <- factor(animalsize)

by(legs, size_factor, max)

```

## R basics - Data frames #1

Data frames are another very common data structure in R. Data frames are like tables you might work with in Excel. You can also think of them as a series of vectors of equal length. Importantly, data frames can organize different classes of data for each column (e.g., a column of numerics with a column of factors), which is different from matrices which can only contain one class of data.

```{r dataframes, include = FALSE}

# In R, there are some built in data sets that can help us learn to use code. Create an object with the built in R data frame called mtcars.
CarsData <- mtcars

# To view the whole data frame in Rstudio, use the View() command or click on
# the name of the object in your Global Environment

# There are many ways to import data sets. Sometimes there are packages created by the R community
# with data sets that can be used for teaching. To install a package of data, or a package with new
# functions use the install.packages() function.
# you must use quotations when installing packages
# install.packages('palmerpenguins') 

# To activate a packages functions use the library() command
library(palmerpenguins) # quotations are option when loading package libraries

# To view just the top rows use the head() command. Look at the help information for head() to see
# the different parameters you can use for viewing data sets.
head(penguins)

# Specify the number of rows you want to look at
head(penguins, 3)

# Look at the structure of the data frame with the str() command
str(penguins)

# Access elements in a data frame just like a matrix
penguins[2, 4]
penguins[2, ]

# Get specific items from one column using it's name. Call items from
# a specific column in a data frame with the '$' symbol and indexing.
penguins$island[6:10]

```

### Exercise 8

Using what we learned about grouping with factors, calculate the average 
penguin mass by island.

```{r exercise8}

island_factor <- factor(penguins$island)

by(penguins$body_mass_g, island_factor, mean)

# What is happening here with these results? 
# View() the data set and report possible issues as a note below

# NA's in the body mass values make it impossible for R to calculate
# the mean. Must remove NA values to proceed.

```

## R basics - Data frames #3

Create your own data frame.

```{r creatingdfs}

# Use the data.frame() command to combine vectors of equal length into a data frame called animals
animals <- data.frame(eyes = eyes, legs = legs, ratio = eye_to_leg, class = class_factor)

View(animals)

```

## R basics - Data frames #4

Data can have "wide" and "long" structures, try to use long whenever you can.

```{r pivotingtables} 

# Using the read.table() command you can create data frames from tab-separated data. 
wide <- read.table(textConnection(
"Gene Control Drug1 Drug2 Drug3 Drug4
Actin 20.249 21.352 22.532 23.557 24.555
Tubulin 8.097 8.986 10.058 11.123 12.246"), header = TRUE)

View(wide)

# Pro-tip: read.table() is also very useful for reading in text files. Check out the help tab to find # syntax for read.table(). See also read.csv(), read.csv2(), read.delim(), read.delim2(). 

# Pivot tables and transform the shape of your data with various packages. Install and load the
# package 'reshape2' in order to convert our wide table into a long table.
library(reshape2)

# the melt() function takes our wide data frame and melts the column names into a 
# new character vector.
long <- melt(wide, id.vars = "Gene")
View(long)

# Assign names to the newly created columns.
colnames(long) <- c("Gene", "Treat", "Effect")
View(long)

# R can handle multidimensional data (more than 2D). Many packages are easier to use when 
# your data is in long form.

```


### Exercise 9

Create a new column for fold-increase compared to control

Calculate the max drug effect for each gene (consider making it a factor...)

```{r exercise9, eval=FALSE, include=FALSE}

# On the wide data set
wide[3, ] = NA
wide[3, 2:length(wide)] = wide[1, 2:length(wide)] / wide[1, 2]
wide[3, 1] = "Actin / Control"
max(wide[3, 2:length(wide)])

wide[4, ] = NA
wide[4, 2:length(wide)] = wide[2, 2:length(wide)] / wide[2, 2]
wide[4, 1] = "Tubulin / Control"
max(wide[4, 2:length(wide)])

# On the long data set
ctrlActin = long[long$Gene == "Actin" & long$Treat == "Control", "Effect"]
ctrlTubulin = long[long$Gene == "Tubulin" & long$Treat == "Control", "Effect"]
long$EffectNorm = NA
long[long$Gene == "Actin", "EffectNorm"] = long[long$Gene == "Actin", "Effect"] / ctrlActin
long[long$Gene == "Tubulin", "EffectNorm"] = long[long$Gene == "Tubulin", "Effect"] / ctrlTubulin
max(long[long$Gene == "Actin", "EffectNorm"])
max(long[long$Gene == "Tubulin", "EffectNorm"])

genefactor <- factor(long$Gene)

by(long$EffectNorm, genefactor, max)

```

## R basics - Adding dimensions with lists

Lists can contain a series of objects. While vectors store elements of the same types, lists can include any element or structure in a series. Indexing with lists requires the use of double [[]].

```{r lists}

# Create a named list with a character string, a numeric scalar (one numeric value), and a
# vector of integers from 1 to 3.
my_list <- list(cat = "meow", seven = 7, series = 1:3)
my_list

# Use indexing to call the first object in my_list
my_list[1]

# Use indexing to call all elements of the third object in my_list.
my_list[[3]]

# Use indexing to call the secend element in the third object in my_list.
my_list[[c(3, 2)]]

```

### Exercise 10

Create a list that combines a vector of character strings, a 3x3 matrix, and two scalars
add names to your items

Print the product of the 3rd and 4th elements on your list

```{r exercise10}

practicelist <- list(strings = c("this", "is", "a", "vector", "of", "strings"), 
                     neo = matrix(1:9, byrow = TRUE, nrow = 3),
                     scalar1 = 3,
                     scalar2 = 6)

print(practicelist[[3]]*practicelist[[4]])

```


## R basics - for loops

A loop is super useful way to repeat actions. But often you don't want to repeat the exact same thing, but rather repeat the same action on each element in a list or array. A for loop does that by running the same block (everything between {}) each time a single element changes its value. 

```{r forloops}

# write a for loop that separately prints every object in my_list, no matter the length.
for (i in 1:length(my_list)) {
 print(my_list[[i]])
}

# write a for loop that prints "The year is" for each year between 2010 and 2015.
for (year in 2010:2015) {
 print(paste("The year is", year))
}

```

We might construct a loop where we do not know the number of elements in a list, or the number of elements in your list changes depending on external factors. Use the `length()` function to find out the number of elements in a list, vector, matrix or number of variables in a data frame. 

```{r indexedloop} 

years <- 2010:2015

# create a loop that prints "The year is" for every year in a list of years of any length.
for (i in 1:length(years)) {
 print(paste("The year is", years[i]))
}

```

## R basics - Arrays

An array is a vector with one or more dimensions. A one-dimensional array is basically a vector. A two dimensional array is basically a matrix. Arrays with more than two dimensions are considered n-dimensional matrices. While lists can contain elements of different data types. Arrays contain elements of all the same data type.

```{r arrays}

# Create an array of random numbers with 4 rows and 3 columns: 

# First create a list of 12 random numbers. To use create random numbers use rnorm().
arraydata <- rnorm(12) 

# Assign arraydata to an array with 4 rows and 3 columns
TestArray <- array(data = arraydata, dim = c(4, 3)) 

# Create a loop that creates four different 4x3 arrays with unique random numbers and print them
for(i in 1:4) {
 arraydata <- rnorm(12)
 TestArray <- array(data = arraydata, dim = c(4, 3)) 
 print(TestArray)

}

```

### Exercise 11

Create an array of random numbers with 3 rows and 4 columns. 
Use a double for loop (or 'nested for loop') to print two of these arrays - in what order does it print them? 

```{r exercise11}

for(i in 1:2) {
 arraydata <- rnorm(12)
 TestArray <- array(data = arraydata, dim = c(3, 4)) 
 for(j in i) {
         print(TestArray)
 }

}

```

## R basics - Logicals statements and flow control

Using logical statements allows for the conditional execution of commands. There are a variety of logical statements that can be used to set conditions and checkpoints for downstream analytical procedures.

```{r logicals}

# simple logicals (>, <, > = , < = )
istrue = 5 > 3
istrue

# not (use an ! to get everything except something)
nottrue = !istrue
nottrue

# OR (use || to return TRUE if any given statement is TRUE)
istrue = 5 < 3 || 2 < 3
istrue

# AND (use && to return TRUE if and only if both given statements are TRUE)
isfalse = 5 < 3 && 2 < 3
isfalse

# & vs &&
isfalse = 5 < 3 & 2 < 3
isfalse

a = c(1, 1)
b = c(2, 2)
a>1 && b>1 

```

If/else statements evaluate logical statements to create decision points when writing code. 

```{r flowcontrol}

x = 5
y = 4
if (x<y) {
 print("X < Y")
} else {
 print("Y > = X")
}

```

### Exercise 12 

Write a for loop over the numbers 1-10 and print even/odd statements. 

Hint: the operator `%%` returns the remainder in a modular division statement.

10 / 3 = 3.33333
10 / 3 = 3 remainder 1 (modular division)
10 %% 3 = 1
11 %% 3 = 2

```{r exercise12}

for(i in 1:10) {
        oddeven <- i %% 2
        if (oddeven == 1) {
                print("Odd")
        } else {
                print("Even")
        }
}
```

## R basics - Functions

Functions can be used to perform a specific operation based on your code. Think of your assigned functions as new commands you can use to customize the code you work with. All functions need to have inputs and output (return).

```{r functions}

# Here we will create a function called mymean which calculates the mean of a vector
mymean <- function(myvector) {
 avg <- mean(myvector)
 return(avg)
}

# If we create a list x of 100 random numbers
x <- rnorm(100)
mymean(x) # should return the same value as mean(x)

# we can create an if/else statement that checks if mymean(x) correctly calculates the mean of x
if (mean(x) == mymean(x)) {
 print("this makes sense")
} else {
 print("The world is crazy!") 
}

# Pro-tip: when working with loops and conditional statements, 
# keep track of where you open and close them 
# using indentation (space at the beginning of the line)

```

### Exercise 13

Create a function that gets two inputs, checks that they are numeric scalar and returns their product. 

Hint: look up `is.numeric()` and `length()`.

```{r exercise13}

scalar_checker <- function(item1, item2){
        if (is.numeric(item1) & is.numeric(item2) & length(item1) == 1 & length(item2) == 1) {
                product <- item1*item2
                return(product)
        } else {
                print("Nope, Try Again")
        }
}

scalar_checker(4, 3)

```

### Exercise 14

Update your function to return product, mean, sum of the two inputs. How can you return more than one thing from a function? 

```{r exercise14}

scalar_multireturn <- function(item1, item2){
        if (is.numeric(item1) & is.numeric(item2) & length(item1) == 1 & length(item2) == 1) {
                product <- item1*item2
                avg <- mean(c(item1, item2))
                added <- item1+item2
                returnlist <- list(product = product, mean = avg, sum = added)
                return(returnlist)
        } else {
                print("Nope, Try Again")
        }
}

scalar_multireturn(4, 8)

```
