---
title: "Module 1.3 Plotting"
author: "A. Gamble, I. Ali, N. Zaitlen"
output: 
  html_document:
    toc: TRUE
    toc_float: TRUE
---

## Intro to Plotting
One of the most powerful features of R is it's functionality with data
visualization and plotting. Various developers in the R community have created
robust tools for data visualization and plotting. Some plotting libraries are
immensely extensive, allowing you to create nearly any kind of plot you would
like and more. 

The R developers community has created a variety of packages that are useful
for plotting. Base R has a highly customizable plotting function and is
used on occasion (see plot()). Searching forums is a great way to learn
about the ways you can customize features needed for your own data visualizatoin
approaches. See: https://www.r-bloggers.com/some-basics-for-base-graphics/
More recently, other plotting libraries have been developed and implemented
in various contexts (plotly, leaflet, pheatmap, etc). 

The most commonly used visualization tool in R is ggplot2 which has immense 
customizability and the most extensive documentation and instructional resources 
of the plotting libraries. ggplot2 is based on what is called the 'grammar of graphics'.
This is the idea that you can build nearly every graph from the same components: 
(1) a data set, (2) a coordinate system and (3) 'geoms' which are visual 
marks that represent data points. ggplot2 is able to create a diverse set 
of plots using these three characteristics. Here we will explore how to 
construct a few example plots using ggplot2. For a more exhaustive list 
of ggplot2 functions please review the ggplot2 cheat sheet linked here: 
https://rstudio.com/wp-content/uploads/2015/03/ggplot2-cheatsheet.pdf. 

It is worth fully exploring ggplot2 documentation. There are many 
customizable features for each geometry making ggplot2 one of the most 
comprehensive packages for data visualization in R. 

The art of data visualization requires both an understanding of the data that
you are working with and also an awareness of the important features of the 
data that you want to highlight. Experimenting with different visualization  
strategies can help you better pin down the most informative ways of displaying 
your data. When working with your own data it can be important to look at mean
or median values, the variance and spread, correlations and clusters.

Through our exercises in this session, we aim to highlight some of the more
common plots you might come across in your research. We will cover how to make
these plots using ggplot, or other relevant plotting libraries

## Setup
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(readr) # read data from the web
library(ggplot2) # plotting library
library(ggExtra) # extra features for plotting library
library(dplyr) # data transformation
library(reshape2) # data transformation
library(RColorBrewer) # better color controls for plotting
```

## Data Import

Using the tools and strategies from Module1.1 and Module1.2 we created a data set
to simulate a time-course gene expression experiment. Using read.csv() import the
data to create an object called "dummydata"

In this data set we will observe the expression of 15 genes (measured in counts per
million) over 8 hours. We will use various plotting strategies to understand various
characteristics of this data set.

```{r dummydataimport}

# using the read.csv() command import the GeneExpressionData.csv file
dummydata <- read.csv("GeneExpressionData.csv", header = TRUE, sep = ",")

# notice that the data is in long form.
View(dummydata)

```

## Plotting One Variable

Histograms are some of the most common plots you will see in your research. In this case,
histograms measure the number of values that fall into pre-defined bins. We can use
geom_histogram() to look at the distribution of counts per-million for all the genes
in our data set. 

We can set a fixed number of bins across a data set, or we can set a fixed bin width to
visualize the data.

```{r histogram}

# use geom_histogram to create a histogram to show the gene expression counts
# for all of the data.

# all ggplot commands follow a similar syntax. First within ggplot() you define the
# data set, and the aesthetics (what you want to set as x, y, or z). For histograms
# you only need to inlude one variable within aes()

ggplot(dummydata, aes(MeanCounts)) +
  geom_histogram(bins = 20)

# it looks like our data is a bit skewed. lets set the fill color to change with
# the gene names to see if some genes have higher overall expression than others.
ggplot(dummydata, aes(MeanCounts, fill = GeneNames)) + 
  geom_histogram(bins = 20)

# We can change the width of the bins as well.
ggplot(dummydata, aes(MeanCounts, fill = GeneNames)) + 
  geom_histogram(binwidth = 10)


```

```{r dotplot}
ggplot(dummydata, aes(MeanCounts)) +
  geom_dotplot(dotsize = 0.5, binwidth = 10, method = "histodot")
```


## Adding lines to plot

We can take a look at how gene expression changes over time using line plots.

```{r plottinglines}
# Add a line across the mean of x from the value 1 to 50 on a plot of data x. 
# the rep function allows you to repeat a value a stated number of times
# rep(value, no_of_repetitions)

ggplot(dummydata, aes(TimePoint, MeanCounts, color = GeneNames)) +
  geom_line()

# Filter on one ore more genes in order to view individual genes on the plot
# 
gene15plot <- dummydata %>%
  filter(GeneNames == c("Gene01","Gene02","Gene03","Gene04","Gene05"))

ggplot(gene15plot, aes(TimePoint, MeanCounts, color = GeneNames)) +
  geom_line()

```

Dot plots are another useful way of visualizing each of the points in a trend

```{r addinglayers}
# add dots to your line plots for genes 5-10
gene510plot <- dummydata %>%
  filter(GeneNames == c("Gene06","Gene07","Gene08","Gene09","Gene10"))

ggplot(gene510plot, aes(TimePoint, MeanCounts, color = GeneNames)) +
  geom_line()+
  geom_point()

ggplot(gene510plot, aes(TimePoint, MeanCounts, color = GeneNames)) +
  geom_smooth(method = "auto")+
  geom_point()

# clean up the geom smooth plot by removing the confidence intervals
ggplot(gene510plot, aes(TimePoint, MeanCounts, color = GeneNames)) +
  geom_smooth(method = "auto", se = F)+
  geom_point()

```


```{r plottinglinesfacets}
# separate genes 11 through 15 onto different plots and keep the confidence intervals

ggplot(gene510plot, aes(TimePoint, MeanCounts, color = GeneNames)) +
  geom_smooth(method = "auto") + 
  geom_point()+
  facet_wrap(facets = "GeneNames") 
```

## Boxplots and Customizing Colors
```{r boxplot}
# Create a boxplot showing the distribution of gene expression for genes 11-15
# across all time points.
gene1115plot <- dummydata %>%
  filter(GeneNames == c("Gene11","Gene12","Gene13","Gene14","Gene15"))

ggplot(gene1115plot, aes(x = GeneNames, y = MeanCounts, fill = GeneNames))+
  geom_boxplot()

# Modify the color palette using RColorBrewer
# scale_fill_brewer() modifies the fill of plot objects
# scale_color_brewer() modifies the color of the plot objects
# select the palette from many options which you can find
# https://www.r-graph-gallery.com/38-rcolorbrewers-palettes.html
ggplot(gene1115plot, aes(x = GeneNames, y = MeanCounts, fill = GeneNames))+
  geom_boxplot() +
  scale_fill_brewer(palette = "Blues")

# Customize Axis Titles and Labels
ggplot(gene1115plot, aes(x = GeneNames, y = MeanCounts, fill = GeneNames))+
  geom_boxplot()+
  scale_fill_brewer(palette = "Greens")+
  labs(title = "Box Plots of Gene Counts",
       x = "Gene IDs",
       y = "Counts per Million")

```

RColorBrewer has a variety of preset color palettes and additional features to
expand and modify color palettes as data increases in complexity. As an entry
point, lets take a look at the distribution of gene expression for all genes
in our dummy data set.

## Violin Plots
```{r violinplots}
ggplot(dummydata, aes(x = GeneNames, y = MeanCounts, fill = GeneNames))+
  geom_violin(scale = "width")+
  labs(title = "Violin Plots of Gene Counts",
       x = "Gene IDs",
       y = "Counts per Million")

ggplot(dummydata, aes(x = GeneNames, y = MeanCounts, fill = GeneNames))+
  geom_violin(scale = "width")+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))+
  labs(title = "Violin Plots of Gene Counts",
       x = "Gene IDs",
       y = "Counts per Million")

ggplot(dummydata, aes(x = GeneNames, y = MeanCounts, fill = GeneNames))+
  geom_violin(scale = "width")+
  geom_dotplot(binaxis = "y", stackdir = "center", fill = "black", color = "white", dotsize = 0.7)+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))+
  labs(title = "Violin Plots of Gene Counts",
       x = "Gene IDs",
       y = "Counts per Million")

```

## Modifying Colors with added complexity

```{r complexcolorscales}

# Most color palettes have a fixed number of default colors in their range. In the case of 
# RColorBrewer most color palettes have a range of 9 colors in their default setting. 
# In the case where we have more than 9 categories, we need to expand the palette in order
# to properly color the data. See what happens when you use the default fill settings on our
# violin plots from earlier

ggplot(dummydata, aes(x = GeneNames, y = MeanCounts, fill = GeneNames))+
  geom_violin(scale = "width")+
  geom_dotplot(binaxis = "y", stackdir = "center", fill = "black", color = "white", dotsize = 0.7)+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))+
  scale_fill_brewer(palette = "Blues")+
  labs(title = "Violin Plots of Gene Counts",
       x = "Gene IDs",
       y = "Counts per Million")

# Notice that genes 10-15 have no color. So in this case we need to expand the color palette
# to match the number of variables we are considering. To do this we can count the number of
# unique genes in our data set and tell RColorBrewer to expand the range to that many colors
ncolors <- length(unique(dummydata$GeneNames))

# once we have the number of colors that we need. We can use the scale_fill_manual() command
# to create the color palette needed. Notice the syntax below. Getting the parenthesis
# right is crucial for this code to work.

ggplot(dummydata, aes(x = GeneNames, y = MeanCounts, fill = GeneNames))+
  geom_violin(scale = "width") +
  scale_fill_manual(values = colorRampPalette(brewer.pal(9, "Blues"))(ncolors))+
  geom_dotplot(binaxis = "y", stackdir = "center", fill = "black", color = "white", dotsize = 0.7)+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))+
  labs(title = "Violin Plots of Gene Counts",
       x = "Gene IDs",
       y = "Counts per Million")

```


## Bar Plot

It is important to note that R plots all the values that you give it. What happens if you want to
create a bar plot that investigates the average gene expression for genes 1 through 5

```{r barplot}

# plot the values like we did with he violin and box plots. What do you see?
ggplot(gene15plot, aes(x = GeneNames, y = MeanCounts, fill = GeneNames)) + 
  geom_col() +
  scale_fill_brewer(palette = "Purples")+
  labs(title = "Bar Plots of Average Gene Counts",
       x = "Gene IDs",
       y = "Counts per Million")

# for bar and column plots, you will need to independently calculate the mean/sd for
# each of the genes and then plot them as individual vaules. Use the group_by() and
# summarize() functions to quickly do this.

gene15summary <- gene15plot %>%
  group_by(GeneNames) %>%
  summarise(Mean = mean(MeanCounts),
            StanDev = sd(MeanCounts))

ggplot(gene15summary, aes(x = GeneNames, 
                          y = Mean, 
                          fill = GeneNames, 
                          ymin = Mean-StanDev, 
                          ymax = Mean+StanDev)) +
  geom_col()+
  scale_fill_brewer(palette = "Purples")+
  geom_errorbar(width = 0.5)+
  labs(title = "Bar Plots of Average Gene Counts",
       x = "Gene IDs",
       y = "Counts per Million")

```

Heatmap
```{r heatmap}

ggplot(dummydata, aes(x = as.character(TimePoint), y = GeneNames))+
  geom_tile(aes(fill = MeanCounts))+
  labs(title = "Heatmap of Average Gene Counts",
       x = "Time Point",
       y = "Counts per Million")
```

